% Document template based on LNCS, adapted by Matt Welsh <mdw@cs.berkeley.edu>

% This version is adapted to use PDFTEX to render to PDF directly. 
% If you want to use dvi, you need to change any figures to use '.eps'
% rather than '.pdf', and probably get rid of the hyperref package.

\documentclass{article}
\usepackage{program}
\usepackage{acm-style10} % ACM proceedings formatting
\usepackage{times}       % Use Adobe Times font set
%\usepackage{epsfig,twocolumn}
\usepackage{epsfig}
\usepackage{url}
\usepackage[english]{babel} % mdw: Required to get good hyphenation on RH6.0
                            % (fixed in RH6.1)
\usepackage{subfigure}
\usepackage{graphicx} 
\usepackage{color}
\usepackage[left=1in,top=1in,right=1in,bottom=1in,nohead]{geometry}

\newcommand{\XXXnote}[1]{{\bf\textcolor{red}{XXX: #1}}}
%\newcommand{\XXXnote}[1]{}
\date{}

\begin{document}


%%%%%%%%%%%% THIS IS WHERE WE PUT IN THE TITLE AND AUTHORS %%%%%%%%%%%%

% Title here
\title{\ttlfnt Compression for Wireless Sensor Networks}

% Author names, affiliations, and e-mail below
\author{{\aufnt Daniel Shteremberg} \\ 
{\affaddr Harvard University}\\ 
{\affaddr dshterem@fas.harvard.edu}
\and {\aufnt Jason Waterman} \\
{\affaddr Harvard University}\\ 
{\affaddr waterman@eecs.harvard.edu}
}

\maketitle
%\copyrightspace
\thispagestyle{empty}

%%%%%%%%%%%%%  ABSTRACT GOES HERE %%%%%%%%%%%%%%

\subsection*{Abstract}
\begin{small}
Wireless sensor networks involve collections of many battery operated
nodes that can number in the thousands.  These nodes are extremely
resource constrained, and are often powered by only a pair AA
batteries and are expected to last months or even years on such a
limited power source.  The primary goal of a sensor networks is to
sense and report on their surrounding habitat.  A new class of sensor
network applications have emerged requiring high data rates and high
data fidelity requirements.  Because of the high energy costs to send
and receive data, there can be a large potential energy savings if the
sensor nodes are able to compress their data before transmitting, if a
compression scheme can be made to run under such resource constrained
devices.  In this paper we look at three data sources collected from
actual sensor network deployments: seismic data from volcanos,
acoustic wildlife data, and motion data collected from a body sensor
nework.  We find that traditional compression schemes, in addition to
being difficult to implement on resource constrained nodes, did not
provide very good compression ratios.  What we find is, a variable
block size (VBS) delta encoding, which transmits the offsets of as
many successive samples that fit in a radio packet, takes advantage of
the temporal locality of the measured signals.  In the best case a
compression ratio of 33.1\% is obtained, while other data compresses
in a range of 42.2\% to 62.8\%.  In addition, our VBS delta encoding
has low computational overhead and we have implemented a demonstration
platform that uses our encoding.
\end{small}

%%%%%%%%%%%%%  BODY OF PAPER GOES HERE %%%%%%%%%%%%%%

% Generally you will want to break the paper into separate sections
% and use \input{...} to include them, like so:
\input{intro.tex} 
\input{data.tex}
\input{related.tex}
\input{delta.tex}
\input{experiments.tex}
\input{future.tex}
\input{conclusion.tex}

%%%%%%%%%%%%%  THIS IS WHERE THE BIBLIOGRAPHY GOES %%%%%%%%%

% Change the word "template" below to the name of the .bib file you use
\begin{small}
\bibliographystyle{abbrv} \bibliography{paper}
\end{small}

\end{document}

